---
title: "Downstream Analyses"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
code-block-border-left: true
---

```{r setup, include=FALSE}
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', warning = FALSE, message = FALSE)
```

## Objectives 🎯
- _Identification of spatially-coordinated genes_
- _Identification of spatial domains_
- _Ligand-receptor inference analysis with `CellChat`_

## Overview
Now you will start to explore ways to use **spatial information** to gather conclusions from our sample. You might be thinking of questions like:

- _Are there genes that display a specific pattern of expression?_
- _Are my clusters more or less spatially organized? Can we include spatial information while clustering spots?_
- _What about cell-cell interactions across spots? Are there ways to check them out in spatial data?_

You'll see in this part that most of these questions are readily answered after a few steps given our properly built dataset!

## Spatially-variable genes with `Giotto`
As we hae seen during the introduction, geographical data analysis approaches have been around for decades in the fields of economics and ecology. Ever since spatial profiling technologies were developed, researchers have re-purposed these mathematical approaches to study biological phenomena!
One of the first applications initially developed for _smFISH_ data was to check whether genes/probes displayed a more or less spatially organized pattern in a statistically relevant manner.
Methods like the Moran's _I_ or Getis-Ord _G_ indices are used to explore whether a variable (like gene expression values)

```{r}
gobj <- seuratToGiottoV5(sp, spatial_assay = 'Visium10X')
```

```{r}
# create spatial network
gobj <- createSpatialNetwork(gobject = gobj,
                                method = "kNN", 
                                     k = 6,
                                     maximum_distance_knn = 400,
                                     name = "spatial_network")
```

```{r}
# Plot spatial network
spatPlot2D(gobject = gobj,  
            show_network= TRUE,
            network_color = "blue", 
            spatial_network_name = "spatial_network")
```

```{r}
# Spatially variable genes detection
ranktest <- binSpect(gobj, 
                     bin_method = "rank",
                     calc_hub = TRUE, 
                     hub_min_int = 5,
                     spatial_network_name = "spatial_network")
```

- Inspect the resulting object, what are the top 5 most _spatially variable_ genes?
- Plot them on the sample like shown below (_hint_: use the `spatFeatPlot2D` function)

```{r}
# Plot the top most spatially-variable genes, we need to first re-scale the data after transformation into giotto object
gobj <- normalizeGiotto(gobj)
spatFeatPlot2D(gobj, 
               expression_values = "scaled",
               feats = ranktest$feats[1:3], 
               cow_n_col = 2, 
               point_size = 1)
```

## Check out spatial modules of expression
```{r}
# Cluster the top 500 spatially variable genes into modules
ext_spatial_genes <- ranktest[1:500,]$feats
```

```{r}
# Compute gene-gene spatial correlation
spat_cor_netw_DT <- detectSpatialCorFeats(
    gobj,
    method = "network",
    spatial_network_name = "spatial_network",
    subset_feats = ext_spatial_genes)
```
Cluster the features into 10 modules
```{r}
spat_cor_netw_DT <- clusterSpatialCorFeats(spat_cor_netw_DT, 
                                           name = "spat_netw_clus", 
                                           k = 10)
```

Plot the clusters
```{r}
heatmSpatialCorFeats(gobj,
                     spatCorObject = spat_cor_netw_DT,
                     use_clus_name = "spat_netw_clus",
                     heatmap_legend_param = list(title = NULL))
```

Summarize the activity of modules into metascores at single spot level
```{r}
# Rank clusters (also plots cluster size vs rank dotplot)
netw_ranks <- rankSpatialCorGroups(
  gobj,
  spatCorObject = spat_cor_netw_DT, 
  use_clus_name = "spat_netw_clus")

cluster_genes_DT <- showSpatialCorFeats(spat_cor_netw_DT, 
                                        use_clus_name = "spat_netw_clus", 
                                        show_top_feats = 1)

# Get gene-clusters associations
cluster_genes <- cluster_genes_DT$clus 
names(cluster_genes) <- cluster_genes_DT$feat_ID

# Create metagenes based on modules
gobj <- createMetafeats(gobj, 
                                feat_clusters = cluster_genes, 
                                name = "cluster_metagene")
```

Plot metafeatures

```{r}
spatCellPlot(gobj,
             spat_enr_names = "cluster_metagene",
             cell_annotation_values = netw_ranks$clusters,
             point_size = 1, 
             cow_n_col = 5)
```

- Which are the top 5 genes for each spatial co-expression cluster?

```{r}
# Get a list of the top 30 co-expressed genes per cluster - hide this and have them work it out
coexpr_dt <- data.table::data.table(
  genes = names(spat_cor_netw_DT$cor_clusters$spat_netw_clus),
  cluster = spat_cor_netw_DT$cor_clusters$spat_netw_clus)

data.table::setorder(coexpr_dt, cluster)

top30_coexpr_dt <- coexpr_dt[, head(.SD, 30) , by = cluster]

spatial_genes <- top30_coexpr_dt$genes
```

## HRMF for spatial domains
This is done on the top 30 genes per cluster
```{r}
HMRF_spatial_genes <- doHMRF(gobject = gobj,
                            expression_values = "scaled",
                            spatial_genes = spatial_genes, # This is your previous output! 
                            k = 20,
                            spatial_network_name = "spatial_network",
                            betas = c(0, 10, 5),
                            output_folder = "HMRF_res/")
```

Load up the results and insert them in the spatial object
```{r}
# Add results
gobj <- addHMRF(gobject = gobj, 
                        HMRFoutput = HMRF_spatial_genes,
                        k = 20, 
                        betas_to_add = c(0, 10, 20, 30, 40),
                        hmrf_name = "HMRF")
```

- Display the resulting spatial domains you obtained (_hint_: try using the `spatPlot2D` function)
- Calculate the proportions of the inferred cell types in each domain, does it make sense compared to the morphology of the sample?
- Display cell type proportions across the identified domains using a heatmap (_hint_: see the `pheatmap` package)

## Ligand-receptor activity inference with CellChat
Ever since the widespread generation of scRNA-seq data, computational methods to infer cell-cell interaction events (CCI) have been developed based on co-expression of ligand- and receptor-encoding genes across clusters. Spatial techonologies deepen this analysis by giving us the chance to weight the inferred interaction by their actual spatial proximity, limiting the chance of identifying false positive results.

We first take the original `SeuratObject` and create a `CellChat` object with its own class

```{r}
# Get normalized data matrix
data.input = Seurat::GetAssayData(sp, slot = "data", assay = "Visium10X")

# Source scale factors for um to px conversion
# Of note, the 'spot_diameter_fullres' factor is different from the `spot` in Seurat object and thus users still need to get the value from the original json file. 
scalefactors = jsonlite::fromJSON(txt = file.path("../data/spatial", 'scalefactors_json.json'))

meta = data.frame(labels = Idents(sp), samples = "sample1", row.names = names(Idents(sp))) # manually create a dataframe consisting of the cell labels
meta$samples <- factor(meta$samples)
unique(meta$labels) # check the cell labels

spatial.locs = GetTissueCoordinates(sp, scale = NULL, cols = c("imagerow", "imagecol"))[,c('x','y')] %>% as.matrix()

spot.size = 65 # the theoretical spot size (um) in 10X Visium
conversion.factor = spot.size/scalefactors$spot_diameter_fullres
spatial.factors = data.frame(ratio = conversion.factor, tol = spot.size/2)

d.spatial <- computeCellDistance(coordinates = spatial.locs, ratio = spatial.factors$ratio, tol = spatial.factors$tol)
min(d.spatial[d.spatial!=0]) # this value should approximately equal 100um for 10X Visium data
```

Create a cellchat object
```{r}
# Create CellChat object
cellchat <- createCellChat(object = data.input, meta = meta, group.by = "labels",
                           datatype = "spatial", coordinates = spatial.locs, spatial.factors = spatial.factors)

# Let's see it
cellchat
```

Build a ligand-receptor database
```{r}
CellChatDB <- CellChatDB.human # use CellChatDB.human if running on human data
```

Take a look at it:
```{r}
dplyr::glimpse(CellChatDB$interaction)
```

- How many ligand-receptor pairs in the database refer to secretory signalling pathways? (_hint_: use the `showDatabaseCategory` function from `CellChat`)

Prepare the expression data
```{r}
CellChatDB.use <- subsetDB(CellChatDB, search = "Secreted Signaling", key = "annotation") # use Secreted Signaling

cellchat@DB <- CellChatDB.use

cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
future::plan("multisession", workers = 4) 
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat, variable.both = F)
```

Compute signalling networks
```{r}
# Compute communication probabilities
cellchat <- computeCommunProb(cellchat, type = "truncatedMean", trim = 0.1,
                              distance.use = TRUE, interaction.range = 250, scale.distance = 0.01,
                              contact.dependent = TRUE, contact.range = 100, nboot=10)
# Aggregate results
cellchat <- aggregateNet(cellchat)
```

- Filter results to retain groups with more than 5 spots (_hint_: use the `filterCommunication` function)
```{r, include=FALSE}
cellchat <- filterCommunication(cellchat, min.cells = 5)
```

- Visualize the number of interactions detected and their strenght using _chord diagrams_ (_hint_: use the `netVisual_circle` function)
```{r, echo=FALSE}
par(mfrow = c(1,2), xpd=TRUE)
# Number of interactions
netVisual_circle(cellchat@net$count, vertex.weight = rowSums(cellchat@net$count), weight.scale = T, label.edge= F, title.name = "Number of interactions")

# Interaction strength
netVisual_circle(cellchat@net$weight, vertex.weight = rowSums(cellchat@net$weight), weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```
- Visualize pairwise interactions using a heatmap (_hint_: use the `netVisual_heatmap` function) 
```{r, echo=FALSE}
netVisual_heatmap(cellchat, measure = "count", color.heatmap = "Blues")
```
- Which are the celltypes with the highest level of interactions?
- Which are the top 3 interactions across tumor cells and fibroblasts?

Let's compute the overall activity of CCI at the **pathway level**, in this case the goal is to summarize the interactions that we have retrieved to ones that pertain to specific intracellular pathways in order to get a sense of what is going on in communicating cells!

```{r}
cellchat <- computeCommunProbPathway(cellchat)
```

Let's now see the activity of a specific pathway:

```{r}
pathways.show <- c("IGF") 
# Circle plot
par(mfrow=c(1,1), xpd = TRUE) 
netVisual_aggregate(cellchat, signaling = pathways.show, layout = "circle")
```

- Plot this in space as well like shown below (_hint_: use the `netVisual_aggregate` function)

We can additionally plot a "binarized" sender/receiver version of the plot above for a specific pathway:

```{r}
# Take an input of a ligand-receptor pair and show expression in binary
spatialFeaturePlot(cellchat, pairLR.use = "IGF1_IGF1R", point.size = 1, do.binary = TRUE, cutoff = 0.05, enriched.only = F, color.heatmap = "Reds", direction = 1)
```

## Recap

