---
title: "Help Page"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
code-block-border-left: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', warning = FALSE, message = FALSE)
```

In this page you can find some helpful information in case you get lost in the functionality of Posit or `R`. In any case if you need to troubleshoot anything, please **do not hesitate to raise your hand and ask**!

### The Posit Interface Died
In the case that the Posit interfaces dies, you can follow the instructions on the screen to re-start it. If this happens, **all the variables within your session will disappear therefore functions will break**. If you already saved the results of the analysis in the `results.csv` file (on day 3) and the samples table in the `samples_table.csv` file (on day 1) you can run the following _directly from your `R` console_:

```{r, eval=FALSE}
# Load existing files for results and samples information
res <- read.table("results.csv", sep = ",")
samples <- read.table("samples_table.csv", sep = ",")
```

In this way, you should be able to run all the code starting from the <a href="../docs/pages/day3.html#Saving Results">"saving results" section</a>. In the **case you DID NOT have the files saved**, you will have to re-run everything before the point where your `R` session died, which should not really be a problem if you had _all the code written in a script_ since you can re-run it seamlessly.

### The Posit RAM Is Full
If you are using Posit and you see the `RAM` indicator on the upper right part of your screen <span style="color: red;">turning red</span>, you can try to do the following to solve the problem:

```{r, eval=FALSE}
# Clean up garbage (unused memory) 
gc()

# In the case we need to remove specific variables
myvar <- c(1,2) # A variable
rm(myvar) 
```

